---
title       : Biometry and Beyond
subtitle    : Skills and Best Practices 
author      : Ben Weinstein
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : IR_BLACK      # 
widgets     : [bootstrap]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---

Last Time
---------

> 1. Data Exploration
> 2. ggplot2

---

Writing functions in R
--------------------------
If you have to repeat the same few lines of code more than once, then you really need to write a *function*. Functions are a fundamental building block of R. You use them all the time in R and it's not that much harder to string functions together (or write entirely new ones from scratch) to do more.

* R functions are objects just like anything else. 
* By default, R function arguments are lazy - they're only evaluated if they're actually used:
* Every call on a R object is almost always a function call.

---

Basic components of a function
--------------------------------
* The `body()`, the code inside the function.
* The `formals()`, the "formal" argument list, which controls how you can call the function.
* The `environment()`` which determines how variables referred to inside the function are found.
* `args()` to list arguments.

---

Writing Functions
=================

```{r}
f <- function(x) x
f

formals(f)

environment(f)
```

---

Functions and Environments
==========================

**Question: How do we delete this function from our environment?**

Variables defined inside functions exist in a different environment than the global environment. However, if a function is not defined inside one, it will look one level above.

---

Example.
=========

```{r}
x <- 2
g <- function() { 
  y <- 1
  c(x, y)
}  
g()
```

---

A first useful function.
=========================

```{r}

first <- function(x, y) {
    z <- x + y
    return(z)
}

first(5,7)
```

---

**Try it!**
----------
*  1. Create a function that takes in two arguments, x and y, and computes the x^2 * y
* 2 Create a function that takes in three arguments, and makes a vector from the result
* 3. Create a function that counts the number of matching items. Hint: use %in% to create a logical statement

---

```{r}
add <- function(a, b) {
  return (a + b)
}
vector <- c(3, 4, 5, 6)

add(vector[[1]],vector[[2]])
```

---

What does this function return?
===============================

```{r}
x <- 5
f <- function() {
y <- 10
c(x = x, y = y) }
f()
```

---

## Functions with pre defined values

```{r}
temp <- function(a = 1, b =2) {
    return(a + b)
}

temp()
temp(5,6)
```

---

**Try It!**
----------

* Write a function that takes in a vector and multiplies the sum of the vector by 10. Return a logical statement based on whether the sum is under 1000.
* Write a function that calculates the mean of every column in a dataframe. Have it break gracefully if the column are not numbers, using class(x) == "numeric". Try your function on the iris dataset

---

Functions usually return the last value it computed
====================================================

```{r}
f <- function(x) {
  if (x < 10) {
    0
  } else {
    10
  }
}
f(5)
f(15)
```
---

# Reproducable programming in R
---------------------------------

It's easy to generate reports dynamically in R.

**Basic idea:** Write **data** + **software** + **documentation** (or in this case manuscripts, reports) together.

---

Analysis code is divided into text and code "chunks".
=====================================================

Literate programming involves with three main steps:  

1. Parse the source document and separate code from narratives.
2. Execute source code and return results.
3. Mix results from the source code with the original narratives.

---

Why this is important?
=============================
Results from scientific research have to be reproducible to be trustworthy. We do not want a finding to be merely due to an isolated occurrence. Instill confidence, share data, results and scripts using github.

---
What is markdown?
==================

An incredibly simple semantic file format.Markdown makes it easy for even those without a web-publishing background to write any sort of text (including with links, lists, bullets, etc.) and have it displayed in a variety of formats. 

* [Markdown cheatsheet](https://github.com/adam-p/markdown-here/wiki/Markdown-Cheatsheet)
* [Original markdown reference](http://daringfireball.net/projects/markdown/basics)

---

Creating a basic knitr document
===============================

In RStudio, choose new R Markdown file (easiest way)
or create a new file and save it with extension `.Rmd`.

<pre><code>
```{r}
# some R code
```
</code></pre>

---
Knitr
--------

Hit the Knit HTML button

**What just happened?**

knitr reads the Rmd file, finds and runs the code chunks identified by the backticks, and replaces it with the output of the functions. 

---

Knitr will auto make tables, graphs, and even slides.

Everything in this course was made by typing into Rstudio and using knitr. Period. It is a powerful tool.

---

Version Control and Github
--------------

Make your life easier, safer, and more fun.

* Setup a GitHub account here.
* Download and install Rstudio.
* Download and install the platform-specific version of Git (not GitHub), default options   work well.
* Configure Git with global commands. I have found this step necessary both times I     ran through this process. Open up the bash version of Git and type the following:         

git config –global user.name “your GitHub account name”                                                     git config –global user.email “your email”

* Open Rstudio and set the path to Git executable. Go to Tools > Options > Git/SVN       

---
Make a repo on github.com
==========================

A repository is the location and name for all the files associated with a particular project. The first step is to log into your GitHub account and create a new repository. Make sure you check the box ‘Initialize this repository with a README.’ 

---
Clone your repo
================

* Open Rstudio and go to Project > Create Project > Version Control > Git and paste the url in the github repo

* Now do some work in your new R project and create and save some files. The next step is to ‘commit’ your work – essentially making a copy of all of your script files (i.e., .R files) associated with the R project.

*Commit your work, give it some description

*push to the web.

sleep easy.

---


Thank You
----------

<style>
pre {
  margin-bottom: -10px;
}
</style>
